---
title: "Project 1 Writeup"
author: "Mary-Kathryn Wert and Ian Chang"
output:
  pdf_document:
    latex_engine: xelatex
    toc: no
    number_sections: true
    pandoc_args: ["--extract-media", "."]
  html_document: default
urlcolor: blue
fontsize: 12pt
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>", 
  out.width = "100%"
)
```

# Evaluation of Original Graph
The original graph effectively communicates its main point—that more than half 
of the global area burned by wildfires occurs in Africa—through its strong and 
visually cohesive design. The title is clear and impactful, with the word 
“Africa” highlighted in red to match the corresponding color used for the 
region in the chart, reinforcing the visual emphasis on these results. The use 
of a stacked bar format is appropriate for showing proportions over time and a 
dotted line at 50% subtly underscores the central message highlighted in the 
title of the graph. However, the chart could benefit from several improvements.
A stacked bar chart is appropriate for showing proportions, it is also hard to 
determine the proportions exhibited beyond the base region. Additionally, there
are potential issues with color in terms of accessibility and labeling.

# Visualization of Data
The new visualization created attempts to solve the minor issues we had in the
orginal graph. The new graph is a line graph to which makes it easier to 
determine the specific proportion of wildfire burned for each region per year. 
Along with the representation type, changes were made aesthetically in order 
to make the graph more accessible and easier to understand for the viewer. 
While the original color palette is aesthetically consistent, some of the 
shades can be difficult to distinguish, especially for viewers with impaired 
color vision which is why the new graph uses colors from the brewer color 
palette which is specifically created with friendly colors. Secondly, the 
original chart lacks a proper legend. The use of a legend in the new graph 
addresses providing context for each line while retaining good legibility. 
Grouping North and South America together felt arbitrary as well in the 
labeling of regions against the bar plot, especially when other continents are 
shown separately. Thirdly, the x-axis could be improved by adding more year 
markers. This change is especially important in terms of the choice to make a 
line graph. The markers help the readers quickly understand the temporal scope 
of the data. Although the Americas are often grouped as “New World” regions, 
the ecosystems most affected by wildfire—such as tropical rainforests of 
South America and the boreal forests of North America—are ecologically and 
geographically distinct. Separating them could offer more meaningful insights 
into regional wildfire patterns.


```{r}
install.packages("ggtext")
library(here)
library(tidyverse)
library(ggtext)
library(ggplot2)

wildfire <- read_csv(here("data","annual-area-burnt-by-wildfires.csv"))

wildfire_clean <- wildfire %>%
  filter(Entity %in% c("North America", "South America", "Asia", "Europe", 
                       "Africa", "Oceania")) %>%
  filter(2012 <= Year & 2024 >= Year) %>%
  rename(area_ha = "Annual area burnt by wildfires")

wildfire_clean$Entity <- factor(wildfire_clean$Entity, 
                                levels = c("Europe", 
                                           "Asia", 
                                           "Oceania", 
                                           "North America", 
                                           "South America", "Africa"))

wildfire_clean <- wildfire_clean %>%
  group_by(Year)%>%
  mutate(area_pct = area_ha/sum(area_ha))%>%
  ungroup()

gg_labels <- tibble(
  Entity = c("Europe", "Asia", "Oceania", "North", "and", "South America", 
             "Africa"),
              x = 14,
              y = c(.985, .93, .86, .795, .74, .68, .30), 
  color = c("#BFCBB7", "#C1A686", "#B3A3C2", "#b3c3c5", "#6B6B6B", "#81919c", 
            "#821E1E"))


ggplot(wildfire_clean, aes(x = factor(Year), y = area_pct, fill = Entity)) +
  geom_bar(stat = "identity", width = 0.85) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_x_discrete(breaks = seq(min(wildfire_clean$Year), 
                                max(wildfire_clean$Year), by = 4)) +
  scale_fill_manual(values = c(
    "Africa" = "#821E1E",
    "North America" = "#b3c3c5",
    "South America" = "#81919c",
    "Oceania" = "#B3A3C2",
    "Asia" = "#C1A686",
    "Europe" = "#BFCBB7"
  )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", 
             linewidth = 0.5, alpha = .35) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "black", 
             linewidth = 0.5, alpha = .1) +
  geom_hline(yintercept = 0.75, linetype = "dashed", color = "black", 
             linewidth = 0.5, alpha = .1) +
  labs(
    title = "More than half of the area burned by<br><b>wildfire is in 
    <span style='color:#821E1E'>Africa</span></b>",
    subtitle = "Each region's share of the global area burned by wildfire.",
    x = NULL,
    y = NULL,
    fill = NULL,
    caption = "**Data source:** Global Wildfire Information System (2025)"
  ) +
  geom_text(data = gg_labels, aes(x = x, y = y, label = Entity, hjust = 0, 
                                  color = color), size = 3.5) +
  scale_color_identity() +
  theme_minimal() +
  coord_cartesian(clip = "off") +
  theme(
    legend.position = "none",
    plot.title = ggtext::element_markdown(family = "serif", face = "bold", 
                                          size = 16),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = ggtext::element_markdown(hjust = 0, color = "#6B6B6B"),
    plot.subtitle = element_text(size = 10),
    plot.margin = margin(t = 10, r = 100, b = 10, l = 10, unit = "pt"),
    aspect.ratio = .80
  ) 
```
```{r}
ggplot(wildfire_clean) +
  geom_line(aes(x = Year, y = area_pct, color = Entity)) +
  scale_x_continuous(breaks = seq(min(wildfire_clean$Year), 
                                  max(wildfire_clean$Year), by = 2)) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_color_manual(values = c(
    "Africa" = "#0072B2",
    "North America" = "#E69F00",
    "South America" = "#56B4E9",
    "Oceania" = "#009E73",
    "Asia" = "#F0E442",
    "Europe" = "#CC0000"
  )) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", 
             linewidth = 0.5) +
  labs(
    title = "More than half of the area burned by<br><b>wildfire each year 
    is in <span style='color:#0072B2'>Africa</span></b>",
    subtitle = "Each region's share of the global area burned by wildfire.",
    x = NULL,
    y = "Percentage of Global Area Burned by Wildfire",
    color = "Region",
    caption = "**Data source:** Global Wildfire Information System (2025)"
  ) +
  theme_minimal(base_family = "serif") +
  theme(
    legend.position = "right",
    plot.title = ggtext::element_markdown(family = "serif", face = "bold", 
                                          size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.caption = ggtext::element_markdown(hjust = 0, color = "#6B6B6B"),
    plot.subtitle = element_text(size = 10),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10, unit = "pt")
  )
```
